---
title: "DEGwithGEO"
author: "Implemented by Eunse-Kim"
---


#01 Load package & GEOdata in RData format
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(Biobase)
library(limma)
library(umap)

load("~/3.Project/ASTHMA/SubtypingAsthma/01.DEGwithGEO/GEOdata/GEOdata.RData")
```


#02 Differential Expressed Gene analysis

###1. GSE182503
- Microarray Data
- PBMC
- design (age, PC1, PC2)
- all-female  

####1) Data pre-Processing
```{r}
# List to ExpressionSet
gse182503 <- GSE182503[[1]]

# make proper column names to match toptable 
fvarLabels(gse182503) <- make.names(fvarLabels(gse182503))

# group membership for all samples
gsms <- paste0("00X110100100111000X001100XX100010111X00100X001000X",
               "1100100X1100X1X11000XX1001100110X0100X11X00110X011",
               "00")
sml <- strsplit(gsms, split="")[[1]]

# filter out excluded samples (marked as "X")
sel <- which(sml != "X")
sml <- sml[sel]
gse182503 <- gse182503[ ,sel]

# log2 transformation
ex <- exprs(gse182503)
qx <- as.numeric(quantile(ex, c(0., 0.25, 0.5, 0.75, 0.99, 1.0), na.rm=T))
LogC <- (qx[5] > 100) ||
  (qx[6]-qx[1] > 50 && qx[2] > 0)
if (LogC) { ex[which(ex <= 0)] <- NaN
exprs(gse182503) <- log2(ex) }

```

####2) DEG Analysis Design
```{r}
# assign samples to groups
gs <- factor(sml)
groups <- make.names(c("asthma","healthy"))
levels(gs) <- groups
gse182503$group <- gs
pset <- phenoData(gse182503) #phenotype

# PCA
pca <- prcomp(t(exprs(gse182503)))
#summary(pca)
plot(pca$x[,1:2])

# Matrix design (non-pc)
dataframe <- data.frame(gse182503$group,pset$age,pca$x[,1:2])
colnames(dataframe) <- c('group','age','PC1','PC2')
dataframe$age <- as.numeric(dataframe$age)
design <- model.matrix(~0 + group + age, dataframe)
```

####3) DEG Analysis using Limma
```{r}
# limma fitting (both ori & pc)
fit <- lmFit(gse182503, design)  # fit linear model

# set up contrasts of interest and recalculate model coefficients
cts <- paste('groupasthma', 'grouphealthy', sep="-")
cont.matrix <- makeContrasts(contrasts=cts, levels=design)
fit2 <- contrasts.fit(fit, cont.matrix)

# compute statistics and table of top 250 significant genes
fit2 <- eBayes(fit2, 0.01)
tT <- topTable(fit2, adjust="fdr", sort.by="B", number=50)
tT <- subset(tT, select=c("GENE_NAME","adj.P.Val","P.Value","t","B","logFC"))

#  compute statistics and table of all genes
tT2 <- topTable(fit2, adjust="fdr", sort.by="B", number=Inf)
knitr::kable(tT)

# write.table(tT, file=stdout(), row.names=F, sep="\t")
# write.csv(tT,file="DEG_list.csv")

```

####4) Visualize and quality control test results.

(1) Build histogram of P-values for all genes. Normal test
- assumption is that most genes are not differentially expressed.
```{r}
hist(tT2$adj.P.Val, col = "grey", border = "white", xlab = "P-adj",
     ylab = "Number of genes", main = "P-adj value distribution")
```

(2) Venn diagram of results
```{r}
dT <- decideTests(fit2, adjust.method="fdr", p.value=0.1)
vennDiagram(dT, circle.col=palette())
```

(3)create Q-Q plot for t-statistic
```{r}
t.good <- which(!is.na(fit2$F)) # filter out bad probes
qqt(fit2$t[t.good], fit2$df.total[t.good], main="Moderated t statistic")
```

(4)volcano plot (log P-value vs log fold change)
```{r}
colnames(fit2) # list contrast names
ct <- 1        # choose contrast of interest
volcanoplot(fit2, coef=ct, main=colnames(fit2)[ct], pch=20,
            highlight=length(which(dT[,ct]!=0)), names=rep('+', nrow(fit2)))
```

(5) MD plot (log fold change vs mean log expression)
```{r}
# highlight statistically significant (p-adj < 0.05) probes
plotMD(fit2, column=ct, status=dT[,ct], legend=F, pch=20, cex=1)
abline(h=0)
```

####5) Part.2 General expression data analysis

(1) Expression value distribution
```{r}
par(mar=c(4,4,2,1))
title <- paste ("GSE182503", "/", annotation(gse182503), " value distribution", sep ="")
plotDensities(ex, group=gs, main=title, legend ="topright")
```

(2)) UMAP plot (dimensionality reduction)
```{r}
ex <- na.omit(ex) # eliminate rows with NAs
ex <- ex[!duplicated(ex), ]  # remove duplicates
ump <- umap(t(ex), n_neighbors = 15, random_state = 123)
par(mar=c(3,3,2,6), xpd=TRUE)
plot(ump$layout, main="UMAP plot, nbrs=15", xlab="", ylab="", col=gs, pch=20, cex=1.5)
legend("topright", inset=c(-0.15,0), legend=levels(gs), pch=20,
       col=1:nlevels(gs), title="Group", pt.cex=1.5)
library("maptools")  # point labels without overlaps
pointLabel(ump$layout, labels = rownames(ump$layout), method="SANN", cex=0.6)
```

(3) mean-variance trend, helps to see if precision weights are needed
```{r}
plotSA(fit2, main="Mean variance trend, GSE182503")
```


###2. GSE165934
- Microarray Data
- Whole Blood
- design (sex, PC1, PC2)
- no age

####1) Data pre-Processing
```{r}
# List to ExpressionSet
gse165934 <- GSE165934[[1]]

# make proper column names to match toptable 
fvarLabels(gse165934) <- make.names(fvarLabels(gse165934))

# group membership for all samples
gsms <- "1111111110000000000"
sml <- strsplit(gsms, split="")[[1]]

# filter out excluded samples (marked as "X")
sel <- which(sml != "X")
sml <- sml[sel]
gse165934 <- gse165934[ ,sel]

# log2 transformation
ex <- exprs(gse165934)
qx <- as.numeric(quantile(ex, c(0., 0.25, 0.5, 0.75, 0.99, 1.0), na.rm=T))
LogC <- (qx[5] > 100) ||
  (qx[6]-qx[1] > 50 && qx[2] > 0)
if (LogC) { ex[which(ex <= 0)] <- NaN
exprs(gse165934) <- log2(ex) }

```

####2) DEG Analysis Design
```{r}
# assign samples to groups
gs <- factor(sml)
groups <- make.names(c("asthma","healthy"))
levels(gs) <- groups
gse165934$group <- gs
pset <- phenoData(gse165934) #phenotype

# PCA
pca <- prcomp(t(exprs(gse165934)))
summary(pca)
plot(pca$x[,1:2])

# Matrix design (non-pc)
# colnames(pset)
dataframe <- data.frame(gse165934$group,pset$gender,pca$x[,1:2])
colnames(dataframe) <- c('group','sex','PC1','PC2')
dataframe$sex <- as.numeric(dataframe$sex)
design <- model.matrix(~0 + group + sex, dataframe)
```

####3) DEG Analysis using Limma
```{r}
# limma fitting (both ori & pc)
fit <- lmFit(gse165934, design)  # fit linear model

# set up contrasts of interest and recalculate model coefficients
cts <- paste('groupasthma', 'grouphealthy', sep="-")
cont.matrix <- makeContrasts(contrasts=cts, levels=design)
fit2 <- contrasts.fit(fit, cont.matrix)

# compute statistics and table of top 250 significant genes
fit2 <- eBayes(fit2, 0.01)
tT <- topTable(fit2, adjust="fdr", sort.by="B", number=50)
tT <- subset(tT, select=c("SPOT_ID","adj.P.Val","P.Value","t","B","logFC"))
colnames(tT)
#  compute statistics and table of all genes
tT2 <- topTable(fit2, adjust="fdr", sort.by="B", number=Inf)
knitr::kable(tT)

write.table(tT, file=stdout(), row.names=F, sep="\t")
# write.csv(tT,file="DEG_list.csv")
```

####4) Visualize and quality control test results.

(1) Build histogram of P-values for all genes. Normal test
- assumption is that most genes are not differentially expressed.
```{r}
hist(tT2$adj.P.Val, col = "grey", border = "white", xlab = "P-adj",
     ylab = "Number of genes", main = "P-adj value distribution")
```

(2) Venn diagram of results
```{r}
dT <- decideTests(fit2, adjust.method="fdr", p.value=0.1)
vennDiagram(dT, circle.col=palette())
```

(3)create Q-Q plot for t-statistic
```{r}
t.good <- which(!is.na(fit2$F)) # filter out bad probes
qqt(fit2$t[t.good], fit2$df.total[t.good], main="Moderated t statistic")
```

(4)volcano plot (log P-value vs log fold change)
```{r}
colnames(fit2) # list contrast names
ct <- 1        # choose contrast of interest
volcanoplot(fit2, coef=ct, main=colnames(fit2)[ct], pch=20,
            highlight=length(which(dT[,ct]!=0)), names=rep('+', nrow(fit2)))
```

(5) MD plot (log fold change vs mean log expression)
```{r}
# highlight statistically significant (p-adj < 0.1) probes
plotMD(fit2, column=ct, status=dT[,ct], legend=F, pch=20, cex=1)
abline(h=0)
```

####5) Part.2 General expression data analysis

(1) Box-and-Whisker Plot
```{r}
ord <- order(gs)  # order samples by group
palette(c("#1B9E77", "#7570B3", "#E7298A", "#E6AB02", "#D95F02",
          "#66A61E", "#A6761D", "#B32424", "#B324B3", "#666666"))
par(mar=c(7,4,2,1))
title <- paste ("GSE165934", "/", annotation(gse165934), sep ="")
boxplot(ex[,ord], boxwex=0.6, notch=T, main=title, outline=FALSE, las=2, col=gs[ord])
legend("topleft", groups, fill=palette(), bty="n")
```

(2) Expression Value Distribution
```{r}
par(mar=c(4,4,2,1))
title <- paste ("GSE165934", "/", annotation(gse165934), " value distribution", sep ="")
plotDensities(ex, group=gs, main=title, legend ="topright")
```

(3) UMAP plot (dimensionality reduction)
```{r}
ex <- na.omit(ex) # eliminate rows with NAs
ex <- ex[!duplicated(ex), ]  # remove duplicates
ump <- umap(t(ex), n_neighbors = 15, random_state = 123)
par(mar=c(3,3,2,6), xpd=TRUE)
plot(ump$layout, main="UMAP plot, nbrs=15", xlab="", ylab="", col=gs, pch=20, cex=1.5)
legend("topright", inset=c(-0.15,0), legend=levels(gs), pch=20,
       col=1:nlevels(gs), title="Group", pt.cex=1.5)
library("maptools")  # point labels without overlaps
pointLabel(ump$layout, labels = rownames(ump$layout), method="SANN", cex=0.6)
```

(4) Mean-Variance trend, helps to see if precision weights are needed
```{r}
plotSA(fit2, main="Mean variance trend, GSE165934")
```


###3. GSE132006 (Part.1)
- Microarray Data
- Whole Blood
- design (sex, PC1, PC2)
- no age

```{r}

```

###4. GSE132006 (Part.2)
- Microarray Data
- PBMC
- design (sex, PC1, PC2)
- no age

```{r}

```
